---
title: 'UFO Tutorial: Part 2'
author: "Avriana Allen"
date: "11/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plot 4: What are the average lengths of times each shape is seen?

Let's find the average length of time that each shape was seen. **R** has a great function called `mean()` which we can just use. However, we will have to do a little more work than using the `count()` function.

What we want to do, is group all the observations by their shape and then take the average of that number. We're going to use a `group_by()` to collect the data by shape. While this doesn't appear to change the data, it will allow us to use summarize. 

If we wanted to count the number of observations in each shape using this method, we would group by `shape` and summarize using the `n()` function which ennumerates the observations. [(Nothing ever goes in those parentheses)](https://dplyr.tidyverse.org/reference/n.html). It would look like this `ufos %>% group_by(shape) %>% summarize( obs = n())`. But we are looking at the average time. 
This means that we want keep grouping by `shape`, but instead of counting in `summarize()` we want to create a new column that will hold the mean. The variable that hold the data we want is called `duration..seconds.` The observations are in seconds, so we divided it by 60 to have the values in minutes. 
```{r plot4, class.source = NULL}
ufos %>% 
  group_by(shape) %>% 
  summarise(average = mean(duration..seconds.)/60)

```

Let's make a graph. What do you think the code below is doing?

```{r plot4.1}
ufos %>% 
  group_by(shape) %>% 
  summarise(average = mean(duration..seconds.)/60,
            obs = n()) %>% 
  arrange(desc(average)) %>%  
  filter(average <= 20) %>%
  ggplot(aes(shape, average)) +
  geom_bar(aes(fill = obs), stat = "identity") +
  coord_flip()
```



We changed the fill of the bars based on the number of observations. We did this by setting `fill` to obs in aes. If we took it out of `aes()`, it would not based upon `obs`. But if we want we can set it to another color like green or black. While you can use any color hex, **R** has some built in colors. Run `colors()` in the console to see. 


## Plot 5: Where are they being seen?

The next plot is a lot more complicated. We're going to make a map of where the most ufos are seen. Since each observation is recorded with the city where it was seen. We can use that to count the number of observations. But if we're actually going to plot it, we're going to need something more. 

We're going to plot it by mapping each point, via it's `longitude` and `latitude`. Lets return to our friend the `count()` function, but insead of counting for each city, let's also count for each unique `longitude` and `latitude`. In theory, this wil produce the same number of observations as counting just with city, but it appears that there are multiple locations per city. We aren't going to worry about that for now. (While it is possible to fix this issue, it is beyond the scope of this tutorial.)

```{r plot5}
ufos %>% 
  count(city, longitude, latitude)
```


There are way too many observations for us to graph on a map. Let's finish cleaning it and save it as a variable, like we did with the original data set. 

```{r plot 5.1}
ufo_map <- ufos %>% 
  count(city, longitude, latitude) %>% 
  filter(n >= 100)
```

Now that we have the data set we want, we can use it to create a scatterplot.

```{r plot5.2}
ufo_map %>% 
  ggplot(aes(longitude, latitude)) +
  geom_point()
```


This doesn't look that great, which is why we're going to use the [maps package](https://cran.r-project.org/web/packages/maps/maps.pdf) to add some borders using `borders()`.

```{r plot5.3}
ufo_map %>% 
  ggplot(aes(longitude, latitude)) +
  borders("state") +
  coord_quickmap() +
  geom_point() 
```


We can see where our dots are now, but they don't look very beautiful. Let's add a bit more data. We can change the `size` and `color` of our points using the number of observations in the `aes()`. 

```{r plot5.4}
ufo_map %>% 
  ggplot(aes(longitude, latitude)) +
  borders("state") +
  coord_quickmap() +
  geom_point(aes(size = n, color = n))
```

We cleaned the map up a bit more! See if you can figure out wha the extra functions are doing!

```{r plot5.5}
ufo_map %>% 
  ggplot(aes(longitude, latitude)) +
  borders("state") +
  coord_quickmap() +
  geom_point(aes(size = n, color = n), alpha = 0.7) +
  theme_void() +
  guides(size = FALSE) + 
  scale_color_gradient(low = "#20dba4", high = "#0c4655") +
  labs(
    x = "",
    y = "",
    color = "Totals"
  ) +
  theme(
  legend.box.margin = margin(0,10,0,0)
  )
```
